Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-06 Manuscript/TEM_draft.tex
Edits to discussion

%§7
%Further directions - is that necessary though?
%   -Using real data
One way to address this issue would be to run a similar analysis using real data. The amount of studies using Total Evidence data type matrices is steadily increasing \citep[e.g.][]{ronquista2012,slaterphylogenetic2013,beckancient2014} and one could analysis these total evidence matrices in the same framework as this study. However, the true topology (obtained from a matrix with no missing data) is unlikely to be known. A second way to address this issue, and especially to incorporate the variation of signal in morphological character would be to simulate morphological characters at different rates \citep{wrightbayesian2014} or to apply patterns of missing data directly obtained from the fossil record \citep{pattinsonphylogeny2014}. Nevertheless, our simulation already show clear results of the effect of missing data on topology in a Total Evidence framework. Also, our results show how recovering the correct topology can be improved by using Bayesian consensus trees in combination with a maximum amount of characters and a of living taxa with morphological data.
%   -Looking at branch length?

@@

%§8
%Back to the importance of TEM - How we can use fossils (Intro-§2)
%   -Data is not a major problem
Despite the various caveats discussed above, we answered our main question on how does Total Evidence matrices perform in recovering topology with different methods and different amounts of missing data.
%   -Good and promising method.
Our results shows that this method performance decreases mainly with a reduction in the overlap of morphological data. In order to address this eventual problem, we propose to increase the data overlap by coding living species and to have a large morphological matrix. Also, our results shows disparity among methods and a clear superiority of the Bayesian consensus tree topology over the Maximum Likelihood one.

@@

%§Conclusions
% part 1- Fossils are important and can be used combined with living taxa (Intro-§1 & §2)
%   -TEM works with missing data
% part 2- This is how we can improve the data to use them in the best way possible (Conclusion-§1 & §8)
%   -Use bayesian consensus
%   -Go code living taxa
Using Total Evidence matrices with missing data is not a problem for recovering the correct topology as long as enough morphological data is overlapping between the fossil and living taxa. In order to optimize good topological recovery, we propose to use the Bayesian consensus tree over the Maximum Likelihood tree or over a random subset of the Bayesian posterior trees distribution. Also, in order to improve the overlap in data in the morphological part of the matrix, we advice to use all the morphological characters available and to code them for a maximum of living taxa present in the matrix.

%---------------------------------------------
%
%       ...
%
%---------------------------------------------
\section{Supplementary material}
Supplementary material (code, analysis and full results) can be found in the Dryad data repository at http://dx.doi.org/10.5061/dryad.XXXX. %section and sentence format asked by Syst. Biol.

\section{Acknowledgements}
Thanks to Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Andrew Jackson, Gavin Thomas and April Wright. %TG : Not sure how many people we have to thank but it could also include Graham Slater, Nick Matzke and Trevor Hodkinson.
for useful comments on our simulation protocol.
%Thanks to Sive Finlay, Kevin Healy and Adam Kane for comments on the manuscript. %TG: or helping writing sentences does not really count.
Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

@@

\bibliographystyle{sysbio}
\bibliography{References}
%Needs to be cleaned a bit

%---------------------------------------------
%
%       Supplementary, separated
%
%---------------------------------------------

>>>>

%§7
%Further directions - is that necessary though?
% NC: Cut for now, might want to put these ideas in the appropriate places above instead.
%   -Using real data
%One way to address this issue 
% NC: Which issue???
%would be to run a similar analysis using real data. The amount of studies using Total Evidence data type matrices is steadily increasing \citep[e.g.][]{ronquista2012,slaterphylogenetic2013,beckancient2014} and one could analysis these total evidence matrices in the same framework as this study. However, the true topology (obtained from a matrix with no missing data) is unlikely to be known. A second way to address this issue, and especially to incorporate the variation of signal in morphological character would be to simulate morphological characters at different rates \citep{wrightbayesian2014} or to apply patterns of missing data directly obtained from the fossil record \citep{pattinsonphylogeny2014}. Nevertheless, our simulation already show clear results of the effect of missing data on topology in a Total Evidence framework. Also, our results show how recovering the correct topology can be improved by using Bayesian consensus trees in combination with a maximum amount of characters and a of living taxa with morphological data.
%   -Looking at branch length?

@@

%§8
%Back to the importance of TEM - How we can use fossils (Intro-§2)
%   -Data is not a major problem
%Despite the various caveats discussed above, we answered our main question on how does Total Evidence matrices perform in recovering topology with different methods and different amounts of missing data.
% NC: For a concluding paragraph, don't start by reminding people of the caveats! End strongly and confidently
% Also no need to restate the question - instead give the answer.
%   -Good and promising method.
%Our results shows that this method performance decreases mainly with a reduction in the overlap of morphological data. To address this problem, we propose to increase the data overlap by coding living species and to have a large morphological matrix. Also, our results shows disparity among methods and a clear superiority %NC: Not really 
%of the Bayesian consensus tree topology over the Maximum Likelihood one.

@@

%§Conclusions
% part 1- Fossils are important and can be used combined with living taxa (Intro-§1 & §2)
%   -TEM works with missing data
% part 2- This is how we can improve the data to use them in the best way possible (Conclusion-§1 & §8)
%   -Use bayesian consensus
%   -Go code living taxa
Using Total Evidence matrices with missing data is not a problem for recovering the correct topology as long as enough morphological data is overlapping between the fossil and living taxa. %NC: Explain more clearly what you mean by overlap - of characters?
Topological recovery is still best when there is less missing data, therefore we advise filling as many gaps in Total Evidence matrices as possible. Because this is difficult, if not impossible, for fossil data, we recommend coding as many

using as many morphological characters as possible and to code them for a maximum of living taxa present in the matrix.

we propose to use the Bayesian consensus tree over the Maximum Likelihood tree or over a random subset of the Bayesian posterior trees distribution. Also, in order to improve the overlap in data in the morphological part of the matrix,

% NC: Never write "in order to". Just "to" will do

%---------------------------------------------
%
%       ...
%
%---------------------------------------------
\section{Supplementary material}
Supplementary material (R code, analyses, supplementary methods and full results) can be found in the Dryad data repository at http://dx.doi.org/10.5061/dryad.XXXX. %section and sentence format asked by Syst. Biol.

\section{Acknowledgements}
Thanks to Fr\'{e}d\'{e}ric Delsuc, Emmanuel Douzery, Andrew Jackson, Gavin Thomas and April Wright. %TG : Not sure how many people we have to thank but it could also include Graham Slater, Nick Matzke and Trevor Hodkinson.
 % NC: Might be nice to thank Trevor, and it won't add too many lines to thank them all. You can also just calle people A.Wright if you want to reduce word count.
for useful comments on our simulation protocol.
%Thanks to Sive Finlay, Kevin Healy and Adam Kane for comments on the manuscript. %TG: or helping writing sentences does not really count. % NC: Depends how much they helped. I just used to thank the lab for that stuff
Thanks to Paddy Doyle, Graziano D'Innocenzo and Sean McGrath for assistance with the computer cluster. Simulations used the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing and funded through grants from Science Foundation Ireland. This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

@@

\bibliographystyle{sysbio}
\bibliography{References}
%Needs to be cleaned a bit
% NC: Yeah. Leave it til we send it to Gavin but it should be a priority. Almost all of them are wrong! 
%---------------------------------------------
%
%       Supplementary, separated
%
%---------------------------------------------
